---
title: "Friday"
description: "Plotting with ggplot2 and a discussion about AI in bioinformatics"
date: "2026-09-18"
categories: [R, ggplot2, visualisation, AI]
engine: knitr
---

## What we did today

In the morning we learned how to make plots in R with **ggplot2**. In the afternoon we had a discussion about **AI in bioinformatics**.


## Part 1: ggplot2

### Setup

```{r}
#| label: install
#| eval: false
install.packages("ggplot2")   # only needed once
```

```{r}
#| label: setup
#| message: false
library(ggplot2)
```

### The grammar of graphics

Every ggplot is built from the same layers, added with `+`:

| Layer | Question it answers | Example |
|---|---|---|
| **data** | which table? | `ggplot(iris, ...)` |
| **aesthetics** `aes()` | which column goes where (x, y, colour, size...)? | `aes(x = Sepal.Length, colour = Species)` |
| **geoms** | what shape? points, lines, boxes... | `geom_point()`, `geom_boxplot()` |
| **facets** | split into small panels? | `facet_wrap(~ Species)` |
| **scales** | how values map to colours/axes | `scale_colour_manual()`, `scale_y_log10()` |
| **labels** | titles and axis names | `labs(title = ..., x = ...)` |
| **theme** | the look (background, fonts...) | `theme_bw()` |


### 1. A scatter plot

`iris` is a dataset built into R: flower measurements from three species.

```{r}
#| label: fig-first
#| fig-cap: "The simplest ggplot: data + aesthetics + one geom."
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point()
```

### 2. Colour, labels and a theme

```{r}
#| label: fig-colour
#| fig-cap: "Colour mapped to species, with labels and a cleaner theme."
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title = "Petal length vs sepal length",
    x = "Sepal length (cm)",
    y = "Petal length (cm)",
    colour = "Species"
  ) +
  theme_bw()
```

::: {.callout-important}
## Inside or outside `aes()`?
- **Inside** `aes()` = *depends on the data*: `aes(colour = Species)` gives each species its own colour.
- **Outside** `aes()` = *the same for everything*: `geom_point(colour = "blue")` makes all points blue.
:::

### 3. Boxplots with the data points on top

```{r}
#| label: fig-boxplot
#| fig-cap: "A boxplot shows the distribution. Jittered points show every observation."
ggplot(iris, aes(x = Species, y = Petal.Width, fill = Species)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(x = NULL, y = "Petal width (cm)") +
  theme_classic() +
  theme(legend.position = "none")
```

### 4. Facets and trend lines

```{r}
#| label: fig-facets
#| fig-cap: "One panel per species, each with a linear trend line."
#| message: false
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(colour = "steelblue") +
  geom_smooth(method = "lm", colour = "black") +
  facet_wrap(~ Species) +
  theme_bw()
```

### 5. A volcano plot 

A volcano plot is the classic way to show a differential expression result: fold change on the x-axis, significance on the y-axis. The data below is **simulated**, just to practise the plot.

```{r}
#| label: simulate-de
set.seed(42)
n_genes <- 2000
de <- data.frame(
  gene   = paste0("gene_", seq_len(n_genes)),
  log2FC = rnorm(n_genes, mean = 0, sd = 1.2)
)
de$pvalue <- 10^(-abs(de$log2FC) * runif(n_genes, 0.5, 3))
de$direction <- ifelse(de$pvalue < 0.01 & de$log2FC >  1, "Up",
                ifelse(de$pvalue < 0.01 & de$log2FC < -1, "Down", "Not significant"))
head(de)
table(de$direction)
```

```{r}
#| label: fig-volcano
#| fig-cap: "Volcano plot of simulated data. Dashed lines show the cut-offs |log2FC| > 1 and p < 0.01."
ggplot(de, aes(x = log2FC, y = -log10(pvalue), colour = direction)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  scale_colour_manual(values = c("Down" = "#2166ac",
                                 "Not significant" = "grey75",
                                 "Up" = "#b2182b")) +
  labs(x = "log2 fold change", y = "-log10(p-value)", colour = NULL) +
  theme_minimal()
```

### 6. Saving a plot

```{r}
#| label: save
#| eval: false
p <- ggplot(iris, aes(Sepal.Length, Petal.Length, colour = Species)) + geom_point()
ggsave("my_plot.png", plot = p, width = 6, height = 4, dpi = 300)
ggsave("my_plot.pdf", plot = p, width = 6, height = 4)   # vector format, good for papers
```

## Part 2: AI in bioinformatics

The afternoon was a discussion about **AI in bioinformatics**. The main points:

**LLMs predict text, they don't "know" things.** They are trained on huge amounts of internet text, including biased and wrong content. They don't say "I don't know" and can **hallucinate** answers, including citations that don't exist. Every reference an AI gives me must be checked.

**Data privacy.** Sequencing and patient data are sensitive. Pasting data into an online chatbot can send it to a company's servers, depending on the terms of service. A safer habit is to describe the data (column names, data types) instead of uploading it. **Local, open-weight models** keep everything on my own computer, but need good hardware and are usually less capable.

**Learning and productivity.** Relying on AI too much can mean learning less. Studies discussed in the course suggest it does not always make experienced people faster either.

**Responsibility.** I am responsible for everything I use.

**Environmental cost.** Data centres use a lot of electricity and water, so it is worth thinking before sending a query.

**AI and reproducibility.** The habits from this week (version control, clear folder structure, documented decisions) also make it easier for an AI assistant, or a human colleague, to understand and help with a project.

## Cheat sheet

| I want to... | Code |
|---|---|
| start a plot | `ggplot(data, aes(x = ..., y = ...))` |
| scatter / line / bar / boxplot | `geom_point()`, `geom_line()`, `geom_col()`, `geom_boxplot()` |
| colour by a column | `aes(colour = column)` (or `fill =` for areas) |
| one fixed colour | `geom_point(colour = "blue")` |
| split into panels | `facet_wrap(~ column)` |
| add a trend line | `geom_smooth(method = "lm")` |
| set titles and axis names | `labs(title = , x = , y = )` |
| change the look | `theme_bw()`, `theme_minimal()`, `theme(legend.position = "none")` |
| save | `ggsave("plot.png", width = , height = , dpi = 300)` |

## Notes {#notes}

<!-- Add your notes as bullet points below this line, for example:
- something I want to remember
-->

